---
title: 'R tutorial: Network inequity metrics — $\xi_a$ and $\xi_d$'
output:
  pdf_document:
    toc: true
    toc_depth: '2'
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 6, fig.height = 4)
```

This document introduces two closely related ways of measuring how
**inequitably** a resource is distributed across a social network.

## Why a network metric at all?

Existing measures implicitly assumed that all people are equal in their
need for resource access, conflated network position with need, and did
not explicitly allow researchers to define need in measures of resource
access. Furthermore, the usual tools for measuring distributional
unfairness — the Gini coefficient, the Theil index — take a list of values
and ask how unevenly spread they are. They are blind to *who is next to
whom*. Two populations with identical Gini coefficients can be in
completely different situations once you know the social structure: in
one, the people without the resource sit next to people who have it; in
the other, they are six hops away from the nearest holder, on the
periphery of the network.

For anything that travels through social ties — health information, a new
practice, a referral, a piece of advice — that difference is the whole
story. So the metrics here are built around a different primitive:

> **Access is not what you hold. Access is who you can reach that holds a resource.**

A node's access depends on how much resource exists *in its social
network*, discounted by how far away that resource sits. Inequity is then
the shortfall from full access, averaged across the population.

There is a second departure from Gini and Theil worth flagging early.
Those are *equality* measures: they treat all deviations from an even
split the same way. The metrics here are *equity* measures: they carry a
weight vector `w` that lets you say some people's shortfall matters more
than others' — typically because they need the resource more. With
uniform weights you get an equality-flavoured reading; with need-based
weights you get an equity reading. Most of this tutorial uses uniform
weights so that we can isolate the network mechanics, but `w` is where
the equity part of the metric actually lives.

## The two variants

Both metrics answer the same question and differ in exactly one respect:
**how access decays with distance.**

|                        | $\xi_a$ (attenuated)                                        | $\xi_d$ (discrete)                                             |
|------------------------|-------------------------------------------------------------|----------------------------------------------------------------|
| Decay rule             | smooth, exponential                                          | hard boundary                                                   |
| Tuning parameter       | `alpha` — decay rate                                         | `d` — reach in hops                                             |
| Access at distance *d* | `exp(-alpha * d)` — shrinks, never quite zero                | 1 if *d* ≤ cutoff, else 0                                       |
| Reads `R` as           | a continuous quantity                                        | binary (holds it / doesn't)                                     |
| Suits                  | resources that dilute with social distance (information)     | resources with a genuine catchment (an invitation, a referral)  |

Neither is the "correct" one. They are two modelling commitments about
how far a resource travels, and picking between them is a substantive
claim about your case, not a technical detail.

## How this document is organised

1. The shared skeleton behind both metrics (notation and intuition)
2. The two definitions, in code, with a walk-through
3. A toy network small enough to check by hand
4. One resource vector, fed to both metrics
5. Reading the output, node by node
6. Sensitivity: what `alpha` and `d` actually do
7. Need weights: what `w` does, and why it is the equity part
8. Takeaways

Nothing here depends on data you don't have — the whole document runs on
a seven-node path graph you can verify with a pen.

## 0. The shared skeleton

Before any code, it helps to see that both metrics are the same two-step
construction with one interchangeable part.

**Step 1 — access.** For every node *i*, compute a score `A[i]` in [0, 1]
saying how well served that node is. This is where the two metrics differ.

*Attenuated version:* sum the resource held everywhere, each contribution
discounted by its distance from *i*.

$$A_i \;=\; \min\!\Big(1,\; \sum_j R_j \, e^{-\alpha\, d_{ij}}\Big)$$

Note that the sum includes *j = i* itself, where the distance is zero and
so the discount factor is `exp(0) = 1`. A node that holds the resource
counts its own holding in full — as it should.

*Discrete version:* check whether *i* is within `d` hops of anyone
holding the resource.

$$A_i \;=\; \begin{cases} 1 & \text{if } \min_{j:\,R_j > 0} d_{ij} \le d \\ 0 & \text{otherwise} \end{cases}$$

**Step 2 — aggregate the shortfall.** Whatever access rule you used, the
inequity score is the weighted mean of what each node is *missing*:

$$\xi \;=\; \frac{\sum_i w_i\,(1 - A_i)}{\sum_i w_i}$$

So `1 - A[i]` is node *i*'s individual shortfall, and `w[i]` says how
heavily that shortfall should count. Both metrics land on the same scale:

- **ξ = 0** — perfect equity. Everyone the weights care about has full
  access.
- **ξ = 1** — maximal inequity. Nobody has any access (for example,
  nobody holds the resource at all).
- Values in between are the weighted average shortfall, and are
  interpretable as such: ξ = 0.6 means the average person — weighted by
  need — is missing 60% of the access they could have had.

Because it is a mean of per-node shortfalls rather than a summary of a
distribution's shape, ξ is decomposable: you can always ask *which* nodes
are driving the number by looking at the `A` vector the functions return
alongside it.

### Three implementation details worth knowing up front

- **Disconnected components.** `igraph::distances()` returns `Inf` for
  pairs in different components. A node in a component with no resource in
  it gets `A = 0` and therefore a full shortfall of 1. That is the
  intended behaviour — unreachable is unreachable — but it means a
  fragmented network can push ξ up sharply, so it is worth checking
  connectivity before interpreting a high score.
- **The clip at 1.** With `normalize_A = TRUE` (the default), access is
  capped at 1. Without the cap, a node sitting among several holders can
  score above 1 and end up with a *negative* shortfall, quietly
  subsidising other nodes' deficits in the average. Keeping the cap makes
  ξ a genuine average of shortfalls, each bounded in [0, 1].
- **Hop counts, not edge weights.** `distances()` uses the graph's `weight`
  edge attribute if one exists. Both functions pass `weights = NA` to force
  plain hop counts. If you *want* weighted path lengths, that argument is
  the one to change.

```{r libs, message=FALSE}
library(igraph)
```

## 1. The definitions

Both functions take the **same first two arguments**: a graph `G` and a
resource vector `R` — a *named* numeric vector, keyed by node name, with
values in [0, 1] saying how much resource each node holds. They also
share the weight vector `w`, named the same way.

The one input the two treat differently is `R` itself. $\xi_a$ uses it as
a continuous quantity: a node holding 0.3 of the resource contributes 0.3,
discounted by distance. $\xi_d$ reads it as a flag: any node with
`R[node] > 0` is a holder, full stop, and 0.3 is treated no differently
from 1.0.

Feeding a single `R` to both is a deliberate choice. It means any
disagreement between the two numbers comes from the decay rule alone —
smooth taper versus hard cutoff — and not from having quietly handed them
different data.

Read the two function bodies side by side: the aggregation lines at the
bottom are identical, and only the block computing `A` changes.

```{r definitions}
node_keys <- function(G) {
  if (is.null(V(G)$name)) as.character(seq_len(vcount(G))) else as.character(V(G)$name)
}

xi_a <- function(G, R, w = NULL, alpha = 1.0, normalize_A = TRUE) {
  nodes <- node_keys(G)
  if (length(nodes) == 0L) return(list(xi = 1, A = setNames(numeric(0), character(0))))

  if (is.null(w)) w <- setNames(rep(1, length(nodes)), nodes)

  D <- distances(G, weights = NA)          # hop counts; Inf if unreachable
  K <- exp(-alpha * D)
  K[!is.finite(D)] <- 0                    # unreachable contributes nothing

  r <- setNames(rep(0, length(nodes)), nodes)
  shared <- intersect(names(R), nodes)
  r[shared] <- R[shared]

  A <- setNames(as.vector(K %*% r), nodes)
  if (normalize_A) A <- pmin(A, 1)

  total_weight <- sum(w)
  if (total_weight == 0) return(list(xi = 1, A = A))

  xi <- sum(w[nodes] * (1 - A)) / total_weight
  list(xi = xi, A = A)
}

xi_d <- function(G, R, w = NULL, d = NULL) {
  nodes <- node_keys(G)
  if (length(nodes) == 0L) return(list(xi = 1, A = setNames(numeric(0), character(0))))

  if (is.null(w)) w <- setNames(rep(1, length(nodes)), nodes)

  total_weight <- sum(w)
  if (total_weight == 0) return(list(xi = 1, A = setNames(numeric(0), character(0))))

  holders <- intersect(names(R)[R > 0], nodes)
  if (is.null(d)) d <- Inf

  A <- setNames(rep(0, length(nodes)), nodes)
  if (length(holders) > 0L) {
    D <- distances(G, v = match(holders, nodes), weights = NA)
    A[apply(is.finite(D) & D <= d, 2, any)] <- 1
  }

  xi <- sum(w[nodes] * (1 - A)) / total_weight
  list(xi = xi, A = A)
}
```

### Walking through $\xi_a$

- `distances(G, weights = NA)` precomputes every pairwise hop count once,
  as a full matrix in vertex order. Unreachable pairs come back `Inf`.
- `K <- exp(-alpha * D)` turns that matrix into discount factors, and
  `A <- K %*% r` is the access formula for every node at once. Writing it
  as a matrix product rather than a double loop is both idiomatic and much
  faster in R. The `K[!is.finite(D)] <- 0` line looks redundant — `exp(-Inf)`
  is already 0 — but it matters when `alpha = 0`, where `-0 * Inf` is `NaN`.
- Rebuilding `r` from `R` against the node list is the defensive step: a
  node missing from `R` contributes 0 rather than erroring, and the
  ordering is guaranteed to line up with `K`'s columns. Note that the
  diagonal of `D` is 0, so a holder's own resource enters at full strength.
- `pmin(A, 1)` applies the cap discussed above.
- The final line is the weighted mean of `1 - A`. The guard on
  `total_weight == 0` returns 1 — the maximally inequitable reading — on
  the grounds that a weighting scheme that cares about nobody cannot
  certify anything as equitable.
- The function returns **both** ξ and the per-node access vector, as a
  list, so you can always drill into the aggregate. R has no tuples; a
  two-element list is the standard substitute.

### Walking through $\xi_d$

- `holders` collects every node with a strictly positive entry in `R`.
- `distances(G, v = holders)` gives a holders × nodes matrix, and
  `apply(..., 2, any)` asks, for each column, whether *any* holder is
  within `d`. That replaces the loop-and-union in the Python version.
- The `is.finite(D) &` guard is not optional: `d` becomes `Inf` when the
  caller passes `NULL`, and `Inf <= Inf` is `TRUE` in R, so without it
  every node in the graph — including unreachable ones — would count as
  served.
- `A` is then a pure 0/1 indicator of membership in that reachable set.
  There is no partial credit: a node one hop past the boundary scores the
  same as a node that is twenty hops away.

## 2. A small toy network

A path graph of 7 nodes makes the effect of distance easy to see by eye.
It is a chain: 0–1–2–3–4–5–6. Node 0 will hold the resource, and because
there is exactly one route between any two nodes, the distance from node 0
to node *i* is simply *i*. That makes every number this document prints
checkable by hand.

igraph is 1-indexed, so vertices would be numbered 1–7 by default. We
assign the names `"0"`–`"6"` explicitly to keep the arithmetic matching
the description above; `node_keys()` then uses those names as the keys for
`R`, `w` and `A`.

Alongside the drawing we also keep the adjacency matrix `adj` — the same
graph as a plain 0/1 matrix. It isn't used by the metrics (they work off
the igraph object directly), but it is a useful sanity check that the
graph is what you think it is: you should see 1s only on the two
off-diagonals.

A path is of course an extreme topology — maximum diameter for its size,
no redundancy, every node a bottleneck. That is precisely why it is a good
first test: effects that are subtle in a realistic network are impossible
to miss here.

```{r toy-graph}
G <- make_ring(7, circular = FALSE)   # a path: 0-1-2-3-4-5-6
V(G)$name <- as.character(0:6)

adj <- as_adjacency_matrix(G, sparse = FALSE)

plot(G, vertex.color = "lightblue", vertex.size = 30,
     main = "Toy network: Path of 7 nodes")

cat("Adjacency matrix (G):\n")
print(adj)
```

## 3. One resource vector, fed to both metrics

Node `0` — one end of the chain — holds the resource, and everyone else
has none. This is the worst-case placement on a path: it maximises the
average distance to the resource, so the ξ values that come out will be
high. Placing it at node 3 instead would halve the longest distance.

`w` gives equal priority weight to every node, so nothing in the
comparison below is driven by weighting choices. This is the *equality*
setting of the metric, and it is the right control for a tutorial about
the decay rules. In substantive use, `w` is where you would encode need —
larger weights on the people for whom lacking the resource matters most —
and the same network can then produce very different ξ values depending on
whether the high-need nodes sit near node 0 or out at node 6.

Note there is only **one** vector here — `R` — used directly by both
$\xi_a$ and $\xi_d$.

```{r resource}
nodes <- node_keys(G)

R <- setNames(as.numeric(nodes == "0"), nodes)
w <- setNames(rep(1, length(nodes)), nodes)

print(R)
```

Colouring the network by who holds the resource makes the setup visible.
Gold is the single holder; everything else is a node whose access has to
come through the chain.

(The layout is chosen by igraph, so the path may not be drawn as a
straight line — the *drawing* is a cosmetic choice, while the distances
that matter are hop counts along the edges.)

```{r resource-plot}
set.seed(0)
plot(G,
     vertex.color = ifelse(R > 0, "gold", "lightblue"),
     vertex.size = 30,
     layout = layout_with_fr(G),
     main = "Toy network colored by resource holding (gold = holds R)")
```

## 4. Compute both metrics on the same `R`

$\xi_a$ is called with `alpha = 1.0`: each additional hop multiplies a
contribution by `exp(-1) ≈ 0.37`, so access falls by roughly two thirds
per step. $\xi_d$ is called with `d = 2`: everything within two hops of
node 0 is fully served, everything beyond is fully excluded.

Both return a list with `$xi` and `$A`, so we can print the aggregate and
then inspect the per-node access scores that produced it. Comparing the
two `A` vectors is the fastest way to internalise what separates the
metrics.

The chunk below is set to `results='asis'`, which is the R Markdown
equivalent of `IPython.display.Markdown` — it sends `cat()` output
straight to the document rather than into a verbatim block, so the `$\xi_a$`
in the strings is rendered as maths rather than printed literally. A plain
`print()` or a default chunk would show the dollar signs.

```{r compute, results='asis'}
res_a <- xi_a(G, R, w = w, alpha = 1.0)
res_d <- xi_d(G, R, w = w, d = 2)

cat(sprintf("$\\xi_a$ ($\\alpha$=1.0) = %.3f\n\n", res_a$xi))
cat(sprintf("$\\xi_d$ ($d$=2) = %.3f\n\n", res_d$xi))

cat("Per-node access under $\\xi_a$ (A):\n\n")
cat(sprintf("- node %s: A=%.3f\n", names(res_a$A), res_a$A), sep = "")

cat("\nPer-node access under $\\xi_d$ (A):\n\n")
cat(sprintf("- node %s: A=%.3f\n", names(res_d$A), res_d$A), sep = "")
```

### Reading the output

**Under $\xi_a$,** the access column is just `exp(-1 * d)` evaluated along
the chain, because there is only one holder:

| node | distance from holder | `A`   |
|------|----------------------|-------|
| 0    | 0                    | 1.000 |
| 1    | 1                    | 0.368 |
| 2    | 2                    | 0.135 |
| 3    | 3                    | 0.050 |
| 4    | 4                    | 0.018 |
| 5    | 5                    | 0.007 |
| 6    | 6                    | 0.002 |

The shortfalls `1 - A` sum to about 5.42, and dividing by 7 gives
**$\xi_a$ ≈ 0.774**. Node 2, two hops out, still carries nonzero access —
the exponential never reaches zero, it only becomes negligible. By node 5
the access is small enough to be arithmetically irrelevant, which is worth
noting: with `alpha = 1`, the metric has effectively stopped
distinguishing between "far" and "very far".

**Under $\xi_d$ with `d = 2`,** the same node 2 counts as *fully*
accessible — it is inside the boundary, so it scores 1.0, exactly like the
holder itself. Node 3 onward scores 0.0. Three of seven nodes are served,
so **$\xi_d$ = 4/7 ≈ 0.571**.

**The two numbers disagree, and that is the point.** 0.774 versus 0.571 on
identical data. The discrete metric is more generous here because it
awards full marks inside its catchment, where the attenuated metric has
already discounted node 1 to 0.37 and node 2 to 0.14. Change `d` to 1 and
the ordering flips. So the two are not interchangeable, and their values
are not directly comparable in level — what *is* comparable is how each one
moves when you change the network, the seeding, or the weights. That is
why the two are typically reported together: agreement in the *pattern*
across conditions is evidence that a result is about structure rather than
about the arbitrary choice of decay rule.

## 5. Sensitivity: `alpha` vs. `d`

`alpha` controls how fast $\xi_a$'s attenuation decays with distance; `d`
controls how far $\xi_d$'s hard boundary extends. They encode the same
intuition — closer to a holder is better — but respond differently.

The two parameters run in opposite directions:

- **Larger `alpha` = steeper decay = higher $\xi_a$.** Distance is punished
  harder, so measured inequity rises. A useful way to calibrate it: access
  halves every `log(2)/alpha ≈ 0.69/alpha` hops. At `alpha = 0.25` that is
  about 2.8 hops, so influence carries a long way; at `alpha = 2` it is
  about a third of a hop, so essentially only immediate neighbours count.
- **Larger `d` = wider catchment = lower $\xi_d$.** More nodes fall inside
  the boundary, so measured inequity falls. `d = NULL` means no bound at
  all — anyone in the same connected component counts as served, which on
  this connected graph drives $\xi_d$ to exactly 0.

Neither parameter is a nuisance knob to be tuned for a nice-looking
result. Each is a claim about the phenomenon: `alpha` says how quickly
influence dissipates through social ties, `d` says how many hops count as
being within reach. Choose them from theory or from what is known about
the diffusion process, fix them, and report the sweep so readers can see
how much of your conclusion rests on the choice.

Note the `list(1, 2, 3, NULL)` rather than `c(...)` for the cutoffs: `NULL`
is a legitimate value of `d` here, and `c()` would silently drop it.

```{r sweep}
alphas  <- c(0.25, 0.5, 1.0, 2.0)
cutoffs <- list(1, 2, 3, NULL)

xi_a_by_alpha  <- sapply(alphas,  function(a) xi_a(G, R, w = w, alpha = a)$xi)
xi_d_by_cutoff <- sapply(cutoffs, function(k) xi_d(G, R, w = w, d = k)$xi)

cat("xi_a by alpha:\n")
cat(sprintf("  alpha=%-5s -> xi_a=%.3f\n", alphas, xi_a_by_alpha), sep = "")

cat("\nxi_d by cutoff:\n")
cat(sprintf("  cutoff=%-5s -> xi_d=%.3f\n",
            sapply(cutoffs, function(k) if (is.null(k)) "NULL" else as.character(k)),
            xi_d_by_cutoff), sep = "")
```

Reading the sweep:

- **$\xi_a$** climbs from 0.466 (`alpha=0.25`) to 0.835 (`alpha=2.0`) —
  monotone increasing, as expected, and the curve flattens at the top.
  Once decay is steep enough that only neighbours count, making it steeper
  changes little, because there is nothing left to take away.
- **$\xi_d$** falls from 0.714 (`cutoff=1`) through 0.571 and 0.429 to
  exactly 0 at `NULL`. On a 7-node path with one holder at the end, each
  extra hop of reach converts exactly one more node from excluded to
  served, so the sequence steps down by 1/7 at a time. The drop to 0 at
  `NULL` is the reminder that an unbounded cutoff makes the metric measure
  connectivity rather than access.
- The ranges overlap: $\xi_a$ ≈ 0.57 sits somewhere around `alpha ≈ 0.4`,
  and $\xi_d$ = 0.571 at `cutoff=2`. You can always find parameter settings
  that make the two agree on one network — but they will not stay agreed
  when the network changes, because they are responding to different
  features of it.

The plot below shows the $\xi_a$ side of that sweep. There is no matching
line for `d` because it is a discrete parameter with one value (`NULL`)
that has no position on a numeric axis.

```{r sweep-plot}
plot(alphas, xi_a_by_alpha, type = "b", pch = 19,
     xlab = expression(alpha), ylab = expression(xi[a]))
```

## 6. Need weights: what `w` does

Everything so far has run with uniform weights, which makes ξ a plain
average of shortfalls. That is an *equality* reading: it treats a hop of
distance as equally bad no matter who is on the far end of it.

The weight vector is what turns the metric into an *equity* measure.
Recall the aggregation step:

$$\xi \;=\; \frac{\sum_i w_i\,(1 - A_i)}{\sum_i w_i}$$

`w[i]` is a claim about how much node *i*'s shortfall should count —
normally standing in for need, vulnerability, or risk. A node with twice
the weight contributes twice as much to the total. Three properties follow
directly from the formula and are worth stating explicitly:

- **Only relative weights matter.** The denominator normalises, so
  multiplying every weight by 100 leaves ξ unchanged. `w` is a profile,
  not a quantity, and you never need to make it sum to anything in
  particular.
- **A zero weight removes a node from the accounting entirely.** Its access
  is still computed, but its shortfall is not counted. That is occasionally
  what you want (a node representing a service rather than a person) and is
  otherwise a strong assumption.
- **`w` never changes `A`.** Weights affect who *counts*, not who *has
  access*. The two questions stay cleanly separated, which is what makes
  the comparison below interpretable.

The important consequence is this: **`G`, `R` and the whole access vector
`A` can be held completely fixed while ξ moves a long way, purely because
of `w`.** Coverage — the share of people served — is blind to that
movement. This is the core claim the metric exists to make.

### 6.1 Need weights on the toy path

Same chain, same holder at node 0, same access scores as before. All that
changes is where we say the need is:

- `w_uniform` — everyone counts equally (the equality baseline).
- `w_near` — the three nodes nearest the holder (0, 1, 2) count three
  times as much as the rest. Need happens to sit where the resource is.
- `w_far` — the three nodes at the far end (4, 5, 6) count three times as
  much. Need sits where the resource is not.

The intervention is identical in all three cases. Only our judgement about
whose exclusion matters has changed.

```{r toy-weights}
idx <- as.integer(nodes)   # "0".."6" back to 0..6 for the comparisons below

w_uniform <- setNames(rep(1, length(nodes)), nodes)                   # equality baseline
w_near    <- setNames(ifelse(idx <= 2, 3, 1), nodes)                  # need near the holder
w_far     <- setNames(ifelse(idx >= 4, 3, 1), nodes)                  # need at the far end

schemes <- list(uniform = w_uniform, near = w_near, far = w_far)

cat(sprintf("%-10s %8s %8s\n", "weights", "xi_a", "xi_d"))
for (nm in names(schemes)) {
  a <- xi_a(G, R, w = schemes[[nm]], alpha = 1.0)$xi
  d <- xi_d(G, R, w = schemes[[nm]], d = 2)$xi
  cat(sprintf("%-10s %8.3f %8.3f\n", nm, a, d))
}

# Only the *relative* weights matter: scaling every weight by 100 changes nothing.
w_near_scaled <- w_near * 100
cat("\nw_near rescaled x100 -> xi_a =",
    round(xi_a(G, R, w = w_near_scaled, alpha = 1.0)$xi, 3), "\n")
```

The spread is substantial, and it comes from nothing but `w`:

| weights | $\xi_a$ | $\xi_d$ |
|---------|---------|---------|
| uniform | 0.774   | 0.571   |
| near    | 0.647   | 0.308   |
| far     | 0.874   | 0.769   |

Under $\xi_d$ the reading swings from 0.308 to 0.769 — from a fairly
equitable-looking intervention to a badly inequitable one — while the set
of served nodes {0, 1, 2} never changes. An evaluator reporting coverage
would report 3/7 in every row and see nothing at all.

$\xi_a$ moves in the same direction but less dramatically (0.647 to 0.874),
which is a general pattern: because it grades access continuously, no node
is ever fully written off, so reweighting cannot swing it as far as it
swings a metric with a hard boundary. The rescaling check at the bottom
confirms the normalisation — `w_near` and `w_near × 100` return exactly the
same value.

### 6.2 Need weights on a network with real structure

The path graph has no degree variation, so "near the holder" and "far from
the holder" had to be imposed by hand. In a scale-free network the
structure supplies the contrast for us: a few hubs with many ties, a long
tail of nodes with two or three.

We build a 100-node Barabási–Albert network and seed the resource at the
3% highest-degree nodes — the standard degree-targeted intervention. Then
we vary only the need profile:

- **`uniform`** — need is unrelated to network position.
- **`degree`** — need is proportional to degree, so the people who need the
  resource most are the well-connected ones.
- **`inv_degree`** — need is proportional to 1/degree, so need is
  concentrated in the periphery.

The middle scheme is the comfortable case for a hub-seeding strategy: it
puts the resource exactly where the need is. The third is the awkward one,
and it is not a contrived scenario — for many health and social outcomes,
isolation and need genuinely travel together.

Two differences from the Python original are worth flagging. First,
igraph's `sample_pa()` and networkx's `barabasi_albert_graph()` are
different generators with different RNGs, so the numbers below will not
match the Python notebook's even at the same nominal parameters — only the
*pattern* should carry over. Second, `sample_pa()` does not guarantee a
minimum degree of `m`: the earliest vertices can end up with degree 1,
which matters for the `inv_degree` weights, so the code checks the minimum
before dividing.

```{r scale-free}
set.seed(7)
SF <- sample_pa(100, m = 2, directed = FALSE)
V(SF)$name <- as.character(seq_len(vcount(SF)))

sf_nodes <- node_keys(SF)
deg <- degree(SF)

# Seed the 3% highest-degree nodes, as a degree-targeted intervention would.
k <- max(1, round(0.03 * vcount(SF)))
seeds <- names(sort(deg, decreasing = TRUE))[seq_len(k)]
R_sf <- setNames(as.numeric(sf_nodes %in% seeds), sf_nodes)

cat(sprintf("nodes: %d, degree range: %d-%d\n", vcount(SF), min(deg), max(deg)))
cat(sprintf("seeded %d nodes: %s (degrees %s)\n",
            k, paste(seeds, collapse = ", "), paste(deg[seeds], collapse = ", ")))
```

```{r scale-free-weights}
w_sf <- list(
  uniform    = setNames(rep(1, length(sf_nodes)), sf_nodes),  # need unrelated to position
  degree     = setNames(as.numeric(deg), sf_nodes),           # need at the centre
  inv_degree = setNames(1 / as.numeric(deg), sf_nodes)        # need at the periphery
)

xi_a_sf <- sapply(w_sf, function(ws) xi_a(SF, R_sf, w = ws, alpha = 1.0)$xi)
xi_d_sf <- sapply(w_sf, function(ws) xi_d(SF, R_sf, w = ws, d = 2)$xi)

# Coverage depends on G and R alone - it cannot see the weights at all.
A_sf <- xi_d(SF, R_sf, w = w_sf$uniform, d = 2)$A
coverage <- mean(A_sf)
cat(sprintf("coverage within 2 hops: %.0f%% (identical under all three schemes)\n\n",
            100 * coverage))

cat(sprintf("%-13s %8s %8s\n", "need weights", "xi_a", "xi_d"))
cat(sprintf("%-13s %8.3f %8.3f\n", names(w_sf), xi_a_sf, xi_d_sf), sep = "")
```

```{r scale-free-plot}
M <- rbind(xi_a_sf, xi_d_sf)
barplot(M, beside = TRUE, names.arg = names(w_sf),
        col = c("steelblue", "darkorange"), ylab = "xi",
        main = "Same network, same seeds - only the need weights differ")
legend("topleft", legend = c(expression(xi[a]~"("*alpha*"=1.0)"), expression(xi[d]~"(d=2)")),
       fill = c("steelblue", "darkorange"), bty = "n")
```

### Reading the scale-free result

**Coverage says the intervention worked.** A large majority of nodes are
within two hops of a seed, from an intervention that touched three nodes.
On that number alone this is a success, and it is the same figure in every
row — coverage is computed from `G` and `R`, so no choice of weights can
move it.

**ξ says it depends on who was missed.** The expected ordering is
`degree` < `uniform` < `inv_degree` under both metrics, with $\xi_d$
showing the wider spread and $\xi_a$ the same ordering on a compressed
scale. Ordering, not magnitude, is what to read across the two metrics.

**Why it moves in that direction.** Hub seeding reaches hubs and their
neighbours first, so the nodes left outside the two-hop boundary have
systematically lower degree than those inside. Weighting by degree
therefore down-weights precisely the people who were missed, and weighting
by inverse degree up-weights them. You can check this directly:

```{r who-was-missed}
missed <- names(A_sf)[A_sf == 0]
cat(sprintf("mean degree, outside 2-hop boundary: %.1f\n", mean(deg[missed])))
cat(sprintf("mean degree, inside 2-hop boundary:  %.1f\n",
            mean(deg[setdiff(names(A_sf), missed)])))

A_cont <- xi_a(SF, R_sf, w = w_sf$uniform, alpha = 1.0)$A
cat(sprintf("mean A (attenuated), degree <= 2: %.2f\n", mean(A_cont[deg <= 2])))
cat(sprintf("mean A (attenuated), degree >= 6: %.2f\n", mean(A_cont[deg >= 6])))
```

**The general lesson.** Adoption and coverage are *effectiveness*
measures, and they are silent about distribution. A degree-targeted
intervention will nearly always look good on them, because hubs are cheap
to reach and reach many others. Whether it looks good on ξ depends on
something coverage cannot represent: the alignment between where the
resource lands and where the need sits. When those are misaligned — the
`inv_degree` row — the intervention is measurably less equitable at
identical coverage, and only a need-weighted metric will say so.

A caveat on this specific demonstration: the gaps here are real but
modest, because a single 100-node draw with three seeds is a small
example. `degree` and `inv_degree` are also fairly gentle need profiles —
sharper ones (say degree² or a threshold on the bottom quartile) separate
the schemes further. Treat the direction as the finding and the size as an
artefact of the toy setup.

## 7. Takeaways

**On the inputs.** $\xi_a$ and $\xi_d$ share the same inputs (`G`, `R`,
`w`). `R` is a continuous resource vector, but for $\xi_d$'s purposes it is
read as binary: `R[node] > 0` means that node holds the resource. Keeping
one `R` for both is what makes the comparison clean.

**On the difference.** $\xi_a$ uses `R`'s magnitude directly as a
distance-discounted quantity (controlled by `alpha`); $\xi_d$ only cares
whether `R` is zero or not, then applies a hard reachability cutoff
(controlled by `d`). Everything else about the two — the shortfall, the
weighting, the aggregation, the [0, 1] scale — is identical.

**On disagreement.** Because of that, $\xi_a$ and $\xi_d$ on the *same* `R`
can disagree in magnitude — that is expected, and reflects the underlying
modelling choice (smooth attenuation vs. hard cutoff), not different data.
Compare them on *patterns across conditions*, not on levels.

**On interpretation.** ξ is a weighted mean of per-node shortfalls, so it
is always decomposable: when a number surprises you, look at the `A` vector
and find the nodes carrying the deficit. A high ξ can mean the resource is
scarce, or that it is badly placed, or that the network is fragmented —
and only the per-node scores tell you which.

**On the weights.** `w` is where the equity claim lives. It changes who
counts, never who has access, and only its relative values matter. Holding
`G`, `R` and `A` fixed and varying `w` alone moved $\xi_d$ from 0.31 to
0.77 on the toy path — all at constant coverage. That gap is exactly what
an effectiveness measure cannot see.

**What is still missing.** Everything here is a single fixed `R`, not a
diffusion process, and one draw of one network rather than a distribution
over runs. The three topologies behave differently under reweighting —
scale-free networks are far more sensitive to it than random or
small-world ones, because they have the degree heterogeneity for need and
position to become misaligned in the first place — and demonstrating that
properly needs repeated runs rather than the single examples used here.
